The Effects of Racial Identity and Discrimination on Political Participation and Political Attitudes

### 2024-12-09 draft_finalreport.qmd
adding nakiyah analysis to draft report

@@

@@

(INSERT PIE CHART)

The model incorporated predictors to examine factors influencing the importance respondents placed on anti-discrimination policies. These included personal experiences with racism, captured by **Q627** (unfair treatment or discrimination) and **Q633** (impact of discrimination on life satisfaction), alongside demographic variables such as **S2_RACE_PRIME** (primary race), **S3** (sexual orientation), **S3B** (gender identity), **S5_AGE** (age), **S13** (education), and **S14** (community type: rural, suburban, urban). Variables **Q629R1–Q629R7** were excluded due to substantial missing data (\>52%). Nonetheless, because it overlapped significantly with **Q627 (see fig. 2)**, the overall essence of experiences with racism was retained in a more general and complete form.

(INSERT ODDS RATIO CHART)

@@

To further address high VIF values observed in predictors such as **Perceptions of Discrimination Against Latinos** (GVIF = 12.45) and **Perceptions of Discrimination Against Immigrants** (GVIF = 11.53), forward and backward stepwise regression was employed. This approach optimized model fit using the Akaike Information Criterion (AIC) while minimizing multicollinearity. The iterative process ensured that only the most significant predictors were retained, reducing VIF values and improving the model's stability.

@@

The confusion matrix reveals that the model has high **sensitivity** (0.9488), thus accurately identifying respondents who consider policies "Not important." However, the low **specificity** (0.1683) indicates that it struggles to correctly classify those who view the policies as "Important." This is expected given the dataset's skew toward "Not important" (74.25%). The **AUC** of 0.7186 suggests moderate predictive ability, but the model's low specificity and imbalance in class distribution highlight areas for improvement.

@@

# Appendix

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,   # Suppress messages
  warning = FALSE    # Suppress warnings
)
```

```{r echo = FALSE}
# Libraries
library(tidyverse)  
library(pROC)
library(vcd)
library(car)
```

```{r echo = FALSE, results = FALSE}
# identifying requireed columns
required_cols <- c(
  # Outcome Variable: Stopping discrimination against racial/ethnic minorities
  "Q1R5",

  # Demographic Variables
  "S2_RACE_PRIME",   # Primary Race
  "S3",              # SexualOrientation
  "S3B",             # Gender
  "S5_AGE",          # Age categories
  "S13",             # Highest level of education
  "S14",             # Community lived in

  # Variables capturing experiences with racism
  "Q629R1",          # Unfair treatment due to racial background/ethnicity
  "Q629R2",          # Unfair treatment due to skin color
  "Q629R3",          # Unfair treatment due to gender
  "Q629R4",          # Unfair treatment due to sexuality/orientation
  "Q629R5",          # Unfair treatment due to immigration status
  "Q629R6",          # Unfair treatment due to religion
  "Q629R7",          # Unfair treatment due to accent

  "Q627",            # Treated unfairly or discriminated against because of your race, ethnicity, gender, sexuality, being an immigrant, religious heritage or having an accent?
  "Q633",            # Discrimination impact on life satisfaction

  # Religion-related variables
  "Q58R1", "Q58R2", "Q58R3", "Q58R4", "Q58R5", "Q58R6", "Q58R7", 
  "Q58R8", "Q58R9", "Q58R10", "Q58R11", "Q58R12", "Q58R13", 
  "Q58R14", "Q58R15", # Various religious affiliations

  # Perceptions of discrimination
  "Q619_Q626R1",     # Perceptions of discrimination against Whites
  "Q619_Q626R2",     # Perceptions of discrimination against Blacks
  "Q619_Q626R3",     # Perceptions of discrimination against Asian Americans
  "Q619_Q626R4",     # Perceptions of discrimination against Native Americans
  "Q619_Q626R5",     # Perceptions of discrimination against Immigrants
  "Q619_Q626R6",     # Perceptions of discrimination against Latinos
  "Q619_Q626R7",     # Perceptions of discrimination against GaysAndLesbians
  "Q619_Q626R8"      # Perceptions of discrimination against Muslims
)

```

```{r echo=FALSE, results=FALSE}
# Create a new dataframe with selected variables
dataset_new <- dataset[, required_cols]

# Recode the outcome variable 'Q1R5' for readability
dataset_new$Q1R5 <- factor(
  ifelse(
    dataset_new$Q1R5 == "(0) NO TO: Stopping discrimination against racial/ethnic minorities", 
    "Not important",
    "Important"  # Remaining cases are "Important"
  ),
  levels = c("Not important", "Important") # Define factor levels
)

# Define the columns for perceptions of discrimination
perception_cols <- c(
  "Q619_Q626R1",  # Perceptions of discrimination against Whites
  "Q619_Q626R2",  # Perceptions of discrimination against Blacks
  "Q619_Q626R3",  # Perceptions of discrimination against Asian Americans
  "Q619_Q626R4",  # Perceptions of discrimination against Native Americans
  "Q619_Q626R5",  # Perceptions of discrimination against Immigrants
  "Q619_Q626R6",  # Perceptions of discrimination against Latinos
  "Q619_Q626R7",  # Perceptions of discrimination against GaysAndLesbians
  "Q619_Q626R8",  # Perceptions of discrimination against Muslims
  "Q633"          # Discrimination impact on life satisfaction
)

# Recode 'Don' as 'Don't know' and ensure columns remain factors
dataset_new[perception_cols] <- lapply(dataset_new[perception_cols], function(x) {
  x <- as.character(x)  # Temporarily convert to character for recoding
  x <- ifelse(x == "(5) Don", "(5) Don't know", x)  # Recode 'Don' to 'Don't know'
  factor(x)  # Convert back to factor
})

# Define required columns to convert to factors (if not already defined)
# Assuming `required_cols` is previously defined; otherwise, specify explicitly
dataset_new[required_cols] <- lapply(dataset_new[required_cols], as.factor)

# Verify the structure of the dataset
str(dataset_new[required_cols])
```

```{r}
# Cleaning and recoding dataset_new
dataset_new <- dataset_new |> 
  mutate(
    # Education recoding
    educ = case_when(
      S13 == "(1) Grades 1-8" ~ "No college experience",
      S13 == "(2) Some High School, but did not graduate" ~ "No college experience",
      S13 == "(3) High School graduate or GED" ~ "No college experience",
      S13 == "(4) Some college" ~ "Some college experience",
      S13 == "(5) Associates, 2-year degree" ~ "College graduate",
      S13 == "(6) Bachelors, 4-year degree" ~ "College graduate",
      S13 == "(7) Post-graduate degree" ~ "College graduate"
    ),
    # Sexual orientation recoding
    sex_orientation = case_when(
      S3 == "(01) Straight, that is, not gay or lesbian" ~ "Straight",
      S3 == "(02) Gay or lesbian" ~ "Queer",
      S3 == "(03) Bisexual" ~ "Queer",
      S3 == "(04) Something else (Specify)" ~ "Queer",
      S3 == "(88) Refused" ~ "Refused/Don't Know",
      S3 == "(99) I don" ~ "Refused/Don't Know"
    ),
    # Urban vs rural recoding
    rural = case_when(
      S14 == "(1) Large urban area" ~ "Urban/suburban",
      S14 == "(2) Large suburb near large city" ~ "Urban/suburban",
      S14 == "(3) Small suburb near small town or city" ~ "Urban/suburban",
      S14 == "(4) Small town or small city" ~ "Urban/suburban",
      S14 == "(5) Rural area" ~ "Rural"
    ),
    # Race recoding
    race_recoded = case_when(
      S2_RACE_PRIME == "(1) White" ~ "White",
      S2_RACE_PRIME == "(2) Hispanic or Latino" ~ "Latino",
      S2_RACE_PRIME == "(3) Black or African American" ~ "Black",
      S2_RACE_PRIME == "(4) Asian American" ~ "Asian American",
      S2_RACE_PRIME == "(5) American Indian/Native American" ~ "AIAN",
      S2_RACE_PRIME == "(6) Arab, Middle Eastern or North African" ~ "Arab/MENA",
      S2_RACE_PRIME == "(7) Native Hawaiian" ~ "NHPI",
      S2_RACE_PRIME == "(8) Not Hawaiian, but other Pacific Islander" ~ "NHPI"
    ),
    # Age recoding
    age_recoded = case_when(
      S5_AGE == "(1) 16-17" ~ "16-17",
      S5_AGE == "(2) 18-29" ~ "18-29",
      S5_AGE == "(3) 30-39" ~ "30-39",
      S5_AGE == "(4) 40-49" ~ "40-49",
      S5_AGE == "(5) 50-59" ~ "50-59",
      S5_AGE == "(6) 60-69" ~ "60-69",
      S5_AGE == "(7) 70 +" ~ "70+"
    )
  ) %>%
  mutate(
    # Converting variables to factors
    race_recoded = factor(race_recoded, levels = c("White", "Black", "Asian American", "Latino", "AIAN", "NHPI", "Arab/MENA")),
    age_recoded = factor(age_recoded, levels = c("18-29", "30-39", "40-49", "50-59", "60-69", "70+")),
    educ = factor(educ, levels = c("No college experience", "Some college experience", "College graduate")),
    rural = factor(rural, levels = c("Rural", "Urban/suburban")),
    sex_orientation = factor(sex_orientation, levels = c("Straight", "Queer", "Refused/Don't Know"))
  )

# Convert individual religion columns to numeric
for (i in 1:15) {
  col_name <- paste0("Q58R", i, "_numeric")
  dataset_new[[col_name]] <- as.numeric(sub("^\\((\\d+)\\).*", "\\1", dataset_new[[paste0("Q58R", i)]]))
}

# Create a new Religion variable
dataset_new$Religion <- "None"

# Assign religion based on the numeric values of each Q58R column
religion_mapping <- c(
  "Q58R1_numeric" = "Catholic", 
  "Q58R2_numeric" = "Protestant", 
  "Q58R3_numeric" = "Other Christian",
  "Q58R4_numeric" = "Jewish", 
  "Q58R5_numeric" = "Muslim", 
  "Q58R6_numeric" = "Hindu", 
  "Q58R7_numeric" = "Buddhist", 
  "Q58R8_numeric" = "Mormon/LDS", 
  "Q58R9_numeric" = "Folk Religion", 
  "Q58R10_numeric" = "Daoist", 
  "Q58R11_numeric" = "Shamanism", 
  "Q58R12_numeric" = "Ancestral Religion", 
  "Q58R13_numeric" = "Spiritual but not religious", 
  "Q58R14_numeric" = "Atheist", 
  "Q58R15_numeric" = "Agnostic"
)

# Loop through the mapping to assign religion values
for (col in names(religion_mapping)) {
  dataset_new$Religion[dataset_new[[col]] == 1] <- religion_mapping[[col]]
}

# Handle multiple religious affiliations
dataset_new$Religion[rowSums(dataset_new[, grep("_numeric", names(dataset_new))] == 1) > 1] <- "Multiple Religions"

# Convert Religion to a factor with specific levels
dataset_new$Religion <- factor(dataset_new$Religion,
                               levels = c("Catholic", "Protestant", "Other Christian", "Jewish", "Muslim",
                                          "Hindu", "Buddhist", "Mormon/LDS", "Folk Religion",
                                          "Daoist", "Shamanism", "Ancestral Religion",
                                          "Spiritual but not religious", "Atheist", "Agnostic", 
                                          "Multiple Religions", "None"))

# Remove redundant religion columns
redundant_cols <- c(
  paste0("Q58R", 1:15), paste0("Q58R", 1:15, "_numeric")
)

# Remove redundant columns from the dataset
dataset_clean <- dataset_new[, !(names(dataset_new) %in% redundant_cols)]

```

```{r echo=FALSE, results=FALSE}
# Check the number of missing values for each variable
missing_summary <- sapply(dataset_clean, function(x) sum(is.na(x)))

# Create a data frame to display the missing values clearly
missing_df <- data.frame(Variable = names(missing_summary), Missing_Count = missing_summary)
print(missing_df)

# Extract the variables Q629R1 to Q629R7
racism_vars <- dataset_clean[, grep("^Q629R[1-7]$", names(dataset_clean))]

# Calculate the proportion of missing values for each variable
missing_proportions <- colSums(is.na(racism_vars)) / nrow(racism_vars)

# Convert to a data frame
missing_summary <- data.frame(Variable = names(missing_proportions),
                              Missing_Proportion = round(missing_proportions, 4))

print(missing_summary)
```

```{r echo=FALSE, results=FALSE}
# Remove all redundant columns
redundant_columns <- c("Q629R1", "Q629R2", "Q629R3", "Q629R4", 
                       "Q629R5", "Q629R6", "Q629R7",
                       "S2_RACE_PRIME", "S13", "S3", "S5_AGE","S14")

dataset_clean <- dataset_clean[, !(names(dataset_clean) %in% redundant_columns)]

head(dataset_clean)
```

```{r echo=FALSE, results=FALSE}
# Rename remaining columns in dataset_clean
colnames(dataset_clean) <- c(
  "ImportanceOfStoppingDiscriminationAgainstMinorities",  # Q1R5
  "Gender",                                               # S3B
  "TreatedUnfairlyOrDiscriminatedAgainst",                # Q627
  "DiscriminationImpactOnLifeSatisfaction",              # Q633
  "PerceptionsOfDiscriminationAgainstWhites",            # Q619_Q626R1
  "PerceptionsOfDiscriminationAgainstBlacks",            # Q619_Q626R2
  "PerceptionsOfDiscriminationAgainstAsianAmericans",    # Q619_Q626R3
  "PerceptionsOfDiscriminationAgainstNativeAmericans",   # Q619_Q626R4
  "PerceptionsOfDiscriminationAgainstImmigrants",        # Q619_Q626R5
  "PerceptionsOfDiscriminationAgainstLatinos",           # Q619_Q626R6
  "PerceptionsOfDiscriminationAgainstGaysAndLesbians",   # Q619_Q626R7
  "PerceptionsOfDiscriminationAgainstMuslims",           # Q619_Q626R8
  "Education",                                           # educ
  "SexOrientation",                                      # sex_orientation
  "Rural",                                               # rural
  "Race",                                                # race_recoded
  "Age",                                                 # age_recoded
  "Religion"                                             # Religion
)

# Verify the renamed columns
colnames(dataset_clean)
```

# Model Building

```{r echo=FALSE, results=FALSE}
# Regression model
model <- glm(ImportanceOfStoppingDiscriminationAgainstMinorities ~ Gender +
               TreatedUnfairlyOrDiscriminatedAgainst + 
               DiscriminationImpactOnLifeSatisfaction +
               PerceptionsOfDiscriminationAgainstWhites + 
               PerceptionsOfDiscriminationAgainstBlacks +
               PerceptionsOfDiscriminationAgainstAsianAmericans + 
               PerceptionsOfDiscriminationAgainstNativeAmericans + 
               PerceptionsOfDiscriminationAgainstImmigrants + 
               PerceptionsOfDiscriminationAgainstLatinos + 
               PerceptionsOfDiscriminationAgainstGaysAndLesbians + 
               PerceptionsOfDiscriminationAgainstMuslims +
               Education + 
                 SexOrientation +
                 Rural +
                 Race +
                 Age +
               Religion,

             data = dataset_clean, family = "binomial")

# Check multicollinearity using VIF
vif(model)
```

## Model 2: stepwise regression
```{r echo=FALSE, results=FALSE}
# Perform stepwise selection
model_stepwise <- step(model, 
                       direction = "both",
                       trace = 0)
```

```{r echo=FALSE, results=FALSE}
# print(model_stepwise)
vif(model_stepwise)
```

```{r echo=FALSE, results=FALSE}

# Extract odds ratios
odds_ratios <- exp(coef(model_stepwise))

# Get confidence intervals for coefficients
conf_intervals <- exp(confint(model_stepwise))

# Extract p-values
p_values <- summary(model_stepwise)$coefficients[, 4]

# Combine everything into a single data frame
results_model_stepwise <- data.frame(
  OddsRatio = odds_ratios,
  LowerCI = conf_intervals[, 1],
  UpperCI = conf_intervals[, 2],
  PValue = p_values
)

# Display results
results_model_stepwise
```

## Model Diagnostics
```{r echo=FALSE, results=FALSE}
# Set seed for reproducibility
set.seed(123)

# Split data into training (70%) and testing (30%)
library(caret)
train_index <- createDataPartition(dataset_clean$ImportanceOfStoppingDiscriminationAgainstMinorities, p = 0.7, list = FALSE)
train_data <- dataset_clean[train_index, ]
test_data <- dataset_clean[-train_index, ]

# Train stepwise logistic regression model on training data

# Fit the initial model (full model)
full_model <- glm(ImportanceOfStoppingDiscriminationAgainstMinorities ~ Gender +
               TreatedUnfairlyOrDiscriminatedAgainst + 
               DiscriminationImpactOnLifeSatisfaction +
               PerceptionsOfDiscriminationAgainstWhites + 
               PerceptionsOfDiscriminationAgainstBlacks +
               PerceptionsOfDiscriminationAgainstAsianAmericans + 
               PerceptionsOfDiscriminationAgainstNativeAmericans + 
               PerceptionsOfDiscriminationAgainstImmigrants + 
               PerceptionsOfDiscriminationAgainstLatinos + 
               PerceptionsOfDiscriminationAgainstGaysAndLesbians + 
               PerceptionsOfDiscriminationAgainstMuslims +
               Education + 
                 SexOrientation +
                 Rural +
                 Race +
                 Age +
               Religion, 
                  data = train_data, family = "binomial")

# Perform stepwise selection
final_model <- step(full_model, 
                       direction = "both",
                       trace = 0)

# Generate predictions on the test data
predicted_probs <- predict(final_model, newdata = test_data, type = "response")

# Convert probabilities to binary outcomes using a threshold (e.g., 0.5)
predicted_classes <- ifelse(predicted_probs > 0.5, "Important", "Not important")

# Convert predicted classes to factor and align levels with actual classes
predicted_classes <- factor(predicted_classes, levels = c("Not important", "Important"))
actual_classes <- factor(test_data$ImportanceOfStoppingDiscriminationAgainstMinorities, 
                                  levels = c("Not important", "Important"))

# Create confusion matrix
conf_matrix <- confusionMatrix(predicted_classes, actual_classes)

# Print the confusion matrix
print(conf_matrix)
```

```{r echo=FALSE, results=FALSE}
# Calculate the ROC curve
roc_curve <- roc(test_data$ImportanceOfStoppingDiscriminationAgainstMinorities, 
                 predicted_probs_stepwise, 
                 levels = c("Not important", "Important"))

# Print the AUC
auc_value <- auc(roc_curve)
cat("AUC:", auc_value, "\n")

# Plot the ROC curve
plot(roc_curve, col = "blue", main = "ROC Curve")
abline(a = 0, b = 1, col = "red", lty = 2)  # Add diagonal line for random guess

# Save the ROC plot
#ggsave("roc_curve_plot.png", width = 8, height = 6)

```

@@

| Response to Q29                   | Count  | Share of Total Respondents |
|-----------------------------------|--------|----------------------------|
| Very interested in politics       | 4,306  | 25%                        |
| Somewhat interested               | 7,612  | 43%                        |
| Not that interested in politics   | 3,643  | 21%                        |
| Not at all interested in politics | 1,984  | 11%                        |
| Total                             | 17,545 | 100%                       |

@@

```{r echo=FALSE}
# Calculate proportions
proportions <- prop.table(table(dataset$Q1R5))
proportions
# Define custom labels
custom_labels <- c("Not important", "Important")

# Combine custom labels with percentages
pie_labels <- paste(custom_labels, round(proportions * 100, 2), "%")

# Plot the pie chart with custom labels
pie(proportions,
    labels = pie_labels,
    main = "Q1R5: Stopping discrimination against racial/ethnic minorities",
    col = c("lightblue", "salmon"))

ggsave("Proportion_Outcome_variable_RQ2.png", width = 20, height = 10)
```

The model incorporated predictors to examine factors influencing the importance respondents placed on anti-discrimination policies. These included personal experiences with racism, captured by Q627 (unfair treatment or discrimination) and Q633 (impact of discrimination on life satisfaction), alongside demographic variables such as S2_RACE_PRIME (primary race), S3 (sexual orientation), S3B (gender identity), S5_AGE (age), S13 (education), and S14 (community type: rural, suburban, urban). Variables Q629R1–Q629R7 were excluded due to substantial missing data (\>52%). Nonetheless, because it overlapped significantly with Q627 (see fig. 2), the overall essence of experiences with racism was retained in a more general and complete form.

```{r echo = FALSE}
# Filter for significant variables (p-value < 0.05)
significant_data <- results_model_stepwise[results_model_stepwise$PValue < 0.05, ]

# Add a column for variable names to use in the plot
significant_data <- significant_data %>% 
  mutate(Variable = rownames(significant_data))

# Create the plot
ggplot(significant_data, aes(x = OddsRatio, y = reorder(Variable, OddsRatio))) +
  geom_point(size = 4, color = "blue") +  # Add points for odds ratios
  geom_errorbarh(aes(xmin = LowerCI, xmax = UpperCI), height = 0.3, color = "black") +  # Add horizontal error bars
  geom_vline(xintercept = 1, linetype = "dashed", color = "red", size = 0.7) +  # Add a vertical reference line at Odds Ratio = 1
  labs(
    title = "Significant Variables with Confidence Intervals (Odds Ratio)",
    x = "Odds Ratio",
    y = "Variables"
  ) +
  theme_minimal() +  # Use a minimal theme for better aesthetics
  theme(
    axis.text.y = element_text(size = 6, color = "black"),  # Improve y-axis text readability
    axis.title.x = element_text(size = 12, face = "bold"),  # Highlight x-axis title
    axis.title.y = element_text(size = 12, face = "bold"),  # Highlight y-axis title
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),  # Center the title
    plot.margin = margin(1, 1, 1, 1, "cm"),  # Adjust margins
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),  # Reduce x-axis text size
    legend.text = element_text(size = 8),  # Reduce legend text size
    legend.title = element_text(size = 9)  # Reduce legend title size
  )

# Save the plot with larger dimensions
ggsave("Significant_Variables_Plot_Enhanced.png", width = 20, height = 10)
```

@@

To further address high VIF values observed in predictors such as Perceptions of Discrimination Against Latinos (GVIF = 12.45) and Perceptions of Discrimination Against Immigrants (GVIF = 11.53), forward and backward stepwise regression was employed. This approach optimized model fit using the Akaike Information Criterion (AIC) while minimizing multicollinearity. The iterative process ensured that only the most significant predictors were retained, reducing VIF values and improving the model's stability.

@@

The confusion matrix reveals that the model has high sensitivity (0.9488), thus accurately identifying respondents who consider policies "Not important." However, the low specificity (0.1683) indicates that it struggles to correctly classify those who view the policies as "Important." This is expected given the dataset's skew toward "Not important" (74.25%). The AUC of 0.7186 suggests moderate predictive ability, but the model's low specificity and imbalance in class distribution highlight areas for improvement.

@@
